Safe Transferable Regions

### 2015-11-18 paper/type-inference.tex
Sec 4. done.

\name's region type system imposes a heavy annotation burden, and the
cost of manually annotating C\# standard libraries with region types
is prohibitive. We now present our region type inference algorithm
that completely eliminates the need to write region type annotations,
except on some higher-order functions. Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., region annotation-erased \FB) to programs in \FB. The key
aspects of this elaboration are described below.

@@

annotation atop the arrow in a method type.}

@@

\input{fb-elabexpr}
\input{fb-elabmeth}

\subsection{Expression Elaboration}

Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
(a). replacing core types occurring in expressions with fresh region
type templates, and (b). explicitly instantiating region parameters
with fresh region variables in method calls and function applications.
This elaboration is performed with respect to the polymorphic type
templates of classes and methods computed as per
\S\ref{sec:fb-templatization}. The elaboration function traverses
expressions in a syntax-directed manner of a type checker, introducing
fresh region type templates for unknown region types, while generating
constraints over region and predicate variables. The precise nature of
generated constraints is explained in \S\ref{sec:fb-constraintsem},
but in summary, they capture the relationships between the type
templates of various subexpressions and their well-formedness.

@@

\subsection{Method and Class Elaboration}

Functions $\elabMeth$ and $\elabClass$ shown in
Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
class definitions, respectively. Both functions first build a context
($\A$) containing a set ($ \rhoenv$) of regions that are currently
live, a map ($\aenv$) mapping type variables to their bounds, and a
constraint formula ($\phicx$) capturing constraints over live region
variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
capture constraints over variables in $\rhoenv$ denoting the fact that
such constraints are yet to be inferred. Both elaboration functions
generate new constraints by checking well-formedness and subtype
relationships among types of their constituents. A fresh constraint
($C_4$) generated by $\elabClass$ captures a couple of safety
conditions: first, the allocation regions of objects referred by the
instance variables should outlive the allocation region of the
instance itself, and second, the allocation regions of a class type
and its superclass type must be the same.

@@

Function $\elabClassTable$ (not shown) elaborates
every definition in the class table $CT$, while accumulating
constraints. The constraints are finally solved for free region and
predicate variables introduced during templatization and elaboration
stages. The solutions are applied to the class table (and to the
artifacts that make up the class table, recursively) to compute a
class table that maps classes to their fully region-annotated
definitions in $\FB$.

@@

Our constraint generation algorithm generates three kinds of
constraints:
\begin{itemize}
\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to the set ($\rhoenv$) of regions in scope,
\item Well-formedness constraints of form
$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas
over region variables ($\rhoenv$) in scope, and
\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
extended with predicate variables and \emph{pending
substitutions}\footnote{we borrowed this terminology
from~\cite{ltpldi08}}:
\begin{smathpar}
\begin{array}{lcl}
\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
    \ALT F(\varphi) \ALT \phi \conj \phi\\
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
\end{itemize}
A pending substitution ($F$) is a substitution function over region
variables/identifiers. They represent the substitutions that need to
be carried out when a predicate variable ($\varphi$) is replaced by a
concrete formula in a validity constraint. For instance, in the
validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.
% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
% In general, validity constraints generated by our algorithm assume one
% of the following two forms:
% \begin{center}
% \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {\phicstr} \qquad
%    \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {F_j(\varphi_j)}
% \)
% \end{center}
% Where $\varphi_i$'s denote the unknown preconditions of the class and
% the method under elaboration. If the constraint is generated while
% checking the well-formedness of a type or elaborating an expression,
% then $\varphi_j$'s denote the unknown preconditions of classes and
% methods that were used in that type or expression. Each use of a
% (region-polymorphic) class or a method has to instantiate region
% parameters, resulting in pending substitutions ($F_j$). Formulas
% $\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
% variables and pending substitutions. While $\phictxt$ captures
% relationships that are \emph{known} to hold between concrete region
% identifiers (i.e., $\rgn$'s) when the constraint was generated,
% $\phicstr$ captures relationships that are \emph{required} to hold
% among region varibles (i.e., $\rho$'s), or relationships between
% region variables and identifiers. Each region variable occuring in
% $\phicstr$ has an associated well-formedness constraint, which
% specifies its unification domain. The unification domain of a
% constraint is the union of unification domains of all region variables
% occuring in the constraint.

@@

Our constraint generation algorithm traverses the entire program,
performing elaboration and collecting constraints. However, we do not
require region type inference to be a whole-program analysis.
Compositionality of type inference is reclaimed by solving the
constraints in a compositional fashion. Concretely, this means that
our constraint solving algorithm visits and solves every constraint
(or, every set of mutually dependent constraints) only once. It
composes computed solutions to solve other constraints that depend on
the solved constraints. Importantly, the failure to solve a dependent
constraint does not result in backtracking. We now describe our
constraint solving algorithm by demonstrating its performance over the
example constraints shown in the previous section. A formal account of
the algorithm can be found in our technical report~\cite{techrep}.
% To simplify the presentation of our algorithm, we assume that are no
% mutually recursive definitions in the source program. Recursive
% definitions are nonetheless allowed.

@@

\paragraph{Example} The dependency DAG ($G_C$) over validity
constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
shown in Fig.~\ref{fig:pair-deps}. Each node (labeled $C_i$) is a set
of constraints that belong to a strongly connected component in the
dependency graph ($G_c$), hence are mutually dependent. All
dependencies, except the self-dependency on $\mathbf{c_{10}}$, are
type-2 dependencies.

A dependency DAG makes the dependencies between constraints explicit.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph. Accordingly, we
obtain a topological ordering of nodes in the graph $G_{{C}}$, and
solve the sets of constraints in that order. The solutions obtained
after solving a constraint set are applied to the constraints in
subsequent sets before attempting to solve them. For the DAG in
Fig.~\ref{fig:pair-deps}, we consider the topological order $[C_1,
C_2, C_3]$, and solve the sets of constraints in that order.

@@

Solving a set ($C$) of constraints involves finding assignments for
free region ($\rho$) and predicate ($\varphi$) variables in the set.
Assignments to region variables can be found via unification (for the
examples in \S~\ref{sec:constraints}, we have already performed the
unification, so $C_{1-3}$ do not contain free region variables). Once
the region variables have been unified, we solve $C$ to find an
assignment for free predicate variables. The first step is to reduce
the set of constraints into an equivalent single constraint by
conjoining antecedents and consequents of all constraints in the
set\footnote{\cite{techrep} formally justifies the soundness of this
step}. For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
below:
\begin{smathpar}
\begin{array}{l}
  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
    {
        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
        \rho_7 = \rho_9 \conj 
    }\\
  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
\end{array}
\end{smathpar}
The constraint $c_{12}$ is a non-recursive constraint of the form
$\isvalid{\phictxt \conj \varphi}{\phicstr}$, where $\phictxt$ and
$\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
this case). Such constraints are solved for $\varphi$ by computing a
maximally weak $\phisol$ such that $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$. If such $\phisol$ satisfies the
well-formedness condition on $\varphi$, then it is clearly a solution
for $\varphi$, because there cannot exist a weaker solution for
$\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
maximally weak $\phisol$ fails the well-formedness condition, then we
establish~\cite{techrep} that it is impossible to satisfy the
constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
and check if it satisfies the well-formedness condition on $\varphi$.
The former problem can be cast as an instance of graph augmentation
problem, as described below.
\vspace*{-0.05in}
\paragraph{Definitions} Given a constraint formula $\phi$, we define
its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
vertices ($V(\phi)$) are free region variables and identifiers in
$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
in\footnote{We alternate between viewing a constraint formula ($\phi$)
as a set of constraints and a conjunction of constraints in this
description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
conjunction of symmetric outlives constraints for the purpose of graph
encoding. Conversely, given a digraph $G$, we define its constraint
encoding $\Phi(G)$ in a straightforward manner. We say that a graph
$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
exist a path between same pair of vertices in $G_1$.

A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
connected as $G(\phicstr)$. The problem of computing such a $G$ is
equivalent to the problem finding of finding minimum number of edges
to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
Efficient algorithms to solve the later problem are known to
exist~\cite{graph}, and we rely on them for the solution.

Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
graph augmentation problems, as described above, results in the
following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
outlives constraints are replaced with equalities):
\begin{center}
\(
  \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1
\)\\
\(
  \varphi_2 \Leftrightarrow \rho_5 = \rho_0
\)
\end{center}
Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
$c_{14}$ gives us a recursive constraint of the following form:
\begin{center}
\(
  \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
\)
\end{center}
Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
$F$ is a substitution function, not necessarily idempotent. Let
$G(\varphi) = \phicstr \conj F(\varphi)$, so that the constraint is
$\isvalid{\phictxt \conj \varphi}{H(\varphi)}$. Our solution to such
recursive constraints  is based on some observations. Firstly, the set
of all possible constraints over a finite set of region
variables/identifiers is a complete meet semi-lattice, where (a).
meet operation is defined as $\phi_1 \sqcap \phi_2 \triangleq \phi_1
\conj \phi_2$, and (b). lattice order is defined as $\phi_1 \le \phi_2
\triangleq \isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Secondly,
function $H$ is a monotone over the lattice since $\forall
\phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow \phi_2)
\Rightarrow (H(\phi_1) \Rightarrow H(\phi_2))}$. From Knaster-Tarski's
theorem, we know that $H$ has a greatest fixed point $\phi_f$ such
that $\isvalid{\cdot}{\phi_f \Leftrightarrow H(\phi_f) }$. To solve
the recursive constraint shown above, we first compute the greatest
fixed point ($\phi_f$) of $H$, and then reduce the constraint to the
following non-recursive constraint:
\begin{center}
\(
  \isvalid{\phictxt \conj \varphi}{\phi_f}
\)
\end{center}
Which can then be solved via the graph-based approach described
before.

Let $H(\varphi_2)$ denote the consequent of the recursive constraint
$c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
\begin{center}
\(
  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
  \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
  \rgn_0 = \rgn_0
\)
\end{center}
Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
\conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
results in the following solution for $\varphi_2$ that meets its
well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
\rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
\begin{center}
\(
  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
  \rho_9 \conj \rho_8 = \rho_9
\)
\end{center}

>>>>

\name's region type system imposes a heavy annotation burden, and the
cost of manually annotating C\# standard libraries with region types
is prohibitive. We now present our region type inference algorithm
that completely eliminates the need to write region type annotations,
except on some higher-order functions. Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions) to programs in \FB. The key aspects of this elaboration
are described below.

@@

annotation atop the arrow.}

@@

%\input{fb-elabexpr}
%\input{fb-elabmeth}

\subsection{Elaboration and Constraint Generation}

Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
(a). replacing core types occurring in expressions with fresh region
type templates, and (b). explicitly instantiating region parameters
with fresh region variables in method calls and function applications.
This elaboration is performed with respect to the polymorphic type
templates of classes and methods computed as per
\S\ref{sec:fb-templatization}. Our expression elaboration
function\footnote{Supplement contains the pseudocode for all the
elaboration functions.} traverses expressions in a syntax-directed
manner of a type checker, introducing fresh region type templates for
unknown region types, while generating constraints over region and
predicate variables. The precise nature of generated constraints is
explained in \S\ref{sec:constraints}, but in summary, they
capture the relationships between the type templates of various
subexpressions and their well-formedness.

@@

Expression elaboration can be trivially lifted to the level of a
method, and method elaboration to the level of a class, while
accumulating constraints. Besides such constraints, elaboration
functions for methods and classes also generate new constraints by
checking well-formedness and subtype relationships among the types of
their constituents. For instance, method elaboration generates a new
constraint that captures the requirement that, under the guarantees
provided by the context, the region type of a method body must be a
subtype\footnote{Subtyping is type equality in \FB. In \name,
subtyping between region types is a straightforward extension of
subtyping between core types.} of method's return type.  Likewise, the
elaboration function for a class generates a fresh constraint that
captures a couple of safety conditions: first, the allocation regions
of objects referred by the instance variables should outlive the
allocation region of the instance itself, and second, the allocation
regions of a class type and its superclass type must be the same.

Elaborating a class table ($CT$) entails elaborating every class
definition in $CT$, while collecting constraints. The constraints thus
collected are finally solved to obtain assignments for free region and
predicate variables introduced during templatization and elaboration
stages. The solutions are applied to the class table (and to the
artifacts that make up the class table, recursively) to compute a
class table that maps classes to their fully region-annotated
definitions in $\FB$.

% \subsection{Method and Class Elaboration}

% Elaborating a method involves performingexpression elaboration over
% its body, while collecting constraints. Method elaboration can be
% lifted to the level of classes

% Functions $\elabMeth$ and $\elabClass$ shown in
% Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
% class definitions, respectively. Both functions first build a context
% ($\A$) containing a set ($ \rhoenv$) of regions that are currently
% live, a map ($\aenv$) mapping type variables to their bounds, and a
% constraint formula ($\phicx$) capturing constraints over live region
% variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
% capture constraints over variables in $\rhoenv$ denoting the fact that
% such constraints are yet to be inferred. Both elaboration functions
% generate new constraints by checking well-formedness and subtype
% relationships among types of their constituents. A fresh constraint
% ($C_4$) generated by $\elabClass$ captures a couple of safety
% conditions: first, the allocation regions of objects referred by the
% instance variables should outlive the allocation region of the
% instance itself, and second, the allocation regions of a class type
% and its superclass type must be the same.

@@

% Function $\elabClassTable$ (not shown) elaborates
% every definition in the class table $CT$, while accumulating
% constraints. The constraints are finally solved for free region and
% predicate variables introduced during templatization and elaboration
% stages. The solutions are applied to the class table (and to the
% artifacts that make up the class table, recursively) to compute a
% class table that maps classes to their fully region-annotated
% definitions in $\FB$.

@@

Three kinds of constraints are generated during elaboration:
\begin{itemize}
\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to the set ($\rhoenv$) of regions in scope,
\item Well-formedness constraints of form
$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas
over region variables ($\rhoenv$) in scope, and
\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
extended with predicate variables and \emph{pending
substitutions}\footnote{we borrowed this terminology
from~\cite{ltpldi08}}:
\begin{smathpar}
\begin{array}{lcl}
\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
    \ALT F(\varphi) \ALT \phi \conj \phi\\
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
\end{itemize}
A pending substitution ($F$) is a substitution function over region
variables/identifiers. They represent the substitutions that need to
be carried out when a predicate variable ($\varphi$) is replaced by a
concrete formula in a validity constraint. For instance, in the
validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.
% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
% In general, validity constraints generated by our algorithm assume one
% of the following two forms:
% \begin{center}
% \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {\phicstr} \qquad
%    \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {F_j(\varphi_j)}
% \)
% \end{center}
% Where $\varphi_i$'s denote the unknown preconditions of the class and
% the method under elaboration. If the constraint is generated while
% checking the well-formedness of a type or elaborating an expression,
% then $\varphi_j$'s denote the unknown preconditions of classes and
% methods that were used in that type or expression. Each use of a
% (region-polymorphic) class or a method has to instantiate region
% parameters, resulting in pending substitutions ($F_j$). Formulas
% $\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
% variables and pending substitutions. While $\phictxt$ captures
% relationships that are \emph{known} to hold between concrete region
% identifiers (i.e., $\rgn$'s) when the constraint was generated,
% $\phicstr$ captures relationships that are \emph{required} to hold
% among region varibles (i.e., $\rho$'s), or relationships between
% region variables and identifiers. Each region variable occuring in
% $\phicstr$ has an associated well-formedness constraint, which
% specifies its unification domain. The unification domain of a
% constraint is the union of unification domains of all region variables
% occuring in the constraint.

@@

Although we collect constraints for the entire program during
elaboration, we solve the constraints in a compositional fashion.
Concretely, this means that our constraint solving algorithm visits
and solves every constraint (or, every set of mutually dependent
constraints) only once. It composes computed solutions to solve other
constraints that depend on the solved constraints. Importantly, the
failure to solve a dependent constraint does not result in
backtracking. We now describe our constraint solving algorithm by
demonstrating its performance over the example constraints shown in
the previous section. A formal account of the algorithm can be found
in the supplement.
% To simplify the presentation of our algorithm, we assume that are no
% mutually recursive definitions in the source program. Recursive
% definitions are nonetheless allowed.

@@

\paragraph{Example} The dependency DAG ($G_C$) over validity
constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
shown below: %in Fig.~\ref{fig:pair-deps}. 
\vspace*{-0.08in}
\begin{figure}[H]
\includegraphics[scale=0.6]{DepDAG.png}
\end{figure}
\vspace*{-0.08in}
Each node (labeled $C_i$) is a set of constraints that belong to a
strongly connected component in the dependency graph ($G_c$), hence
are mutually dependent. All dependencies, except the self-dependency
on $\mathbf{c_{10}}$, are type-2 dependencies.

A dependency DAG makes the dependencies between constraints explicit.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph's transpose.
Accordingly, we obtain a topological ordering of nodes in the graph
$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
in that order. The solutions obtained after solving a constraint set
are applied to the constraints in subsequent sets before attempting to
solve them. For the DAG in figure above, we consider the topological
order $[C_1, C_2, C_3]$ of its transpose, and solve the sets of
constraints in that order.

@@

Solving a set ($C$) of constraints involves finding assignments for
free region ($\rho$) and predicate ($\varphi$) variables in the set.
Assignments to region variables can be found via unification (for the
examples in \S~\ref{sec:constraints}, we have already performed the
unification, so $C_{1-3}$ do not contain free region variables). Once
the region variables have been unified, we solve $C$ to find an
assignment for free predicate variables. The first step is to reduce
the set of constraints into an equivalent single constraint by
conjoining antecedents and consequents of all constraints in the
set\footnote{Supplement contains formal justification for this step}.
For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
below:
\begin{smathpar}
\begin{array}{l}
  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
    {
        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
        \rho_7 = \rho_9 \conj 
    }\\
  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
\end{array}
\end{smathpar}
The constraint $c_{12}$ is a non-recursive constraint of the form
$\isvalid{\phictxt \conj \varphi}{\phicstr}$, where $\phictxt$ and
$\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
this case). Note that the constraint describes an abduction problem,
whose solution is the weakest $\phisol$ such that $\isvalid{\phictxt
\conj \phisol}{\phicstr}$. We solve the problem by casting it as an
instance of graph augmentation problem~\cite{siam92}. Our intuition is
based on the observation that a constraint can be viewed as a directed
graph with region variables as vertices and outlives relationships
between them as edges (equality is encoded as a conjunction of
symmetric outlives relationships).  Let $G(\phictxt)$ and
$G(\phicstr)$ represent such graphs for $\phictxt$ and $\phicstr$,
respectively. Now, $\isvalid{\phictxt}{\phicstr}$ if and only if every
pair $(\rho_1,\rho_2)$ of connected vertices in $G(\phicstr)$ are also
connected in $G(\phictxt)$. If $\rho_1$ and $\rho_2$ are not connected
in $G(\phictxt)$, then it means that $\isvalid{\phicstr}{\rho_1
\outlives \rho_2}$ but $\isnotvalid{\phictxt}{\rho_1 \outlives
\rho_2}$. Since $\phisol$ needs to satisfy $\isvalid{\phictxt \conj
\phisol}{\phicstr}$, it must be the case that $\rho_1$ and $\rho_2$
are connected in $G(\phictxt) \cup G(\phisol)$. Since $\phisol$ must
be weakest such formula, it can be obtained by computing the minimum
number of edges that need to be added to $G(\phictxt)$ such that it is
``as connected as'' $G(\phicstr)$. Using this technique, we obtain the
following solutions to $\varphi_0$ and $\varphi_1$ for constraints in
the running example:
\(
  \varphi_0 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1
\)\\
\(
  \varphi_1 \Leftrightarrow \rho_5 = \rho_0
\)
Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
$c_{14}$ gives us a recursive constraint of the following form:
\(
  \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
\)
Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and $F$ is
a substitution function, not necessarily idempotent. Our solution to the
constraints of the above form is based on the observation that $H(\varphi) =
\phicstr \conj F(\varphi)$ is a monotone over the complete meet semi-lattice
of all possible constraints over the domain of region variables in the
program. Hence, by Tarski's fixpoint theorem~\cite{tarski}, $H$ has a
greatest fixed point. To solve the recursive constraint, we first compute the
greatest fixed point ($\phi_f$) for $H$, and then solve the non-recursive
constraint $\isvalid{\phictxt \conj \varphi}{\phi_f}$ using the graph
approach described above. Using this approach, we solve the recursive
constraint $c_{14}$, to obtain the following solution to $\varphi_2$:
\(
  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
  \rho_9 \conj \rho_8 = \rho_9
\)

% If such $\phisol$ satisfies the
% well-formedness condition on $\varphi$, then it is clearly a solution
% for $\varphi$, because there cannot exist a weaker solution for
% $\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
% maximally weak $\phisol$ fails the well-formedness condition, then we
% establish~\cite{techrep} that it is impossible to satisfy the
% constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
% that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
% and check if it satisfies the well-formedness condition on $\varphi$.

% \vspace*{-0.05in}
% \paragraph{Definitions} Given a constraint formula $\phi$, we define
% its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
% vertices ($V(\phi)$) are free region variables and identifiers in
% $\phi$, and whose edges ($E(\phi)$) denote outlives constraints
% in\footnote{We alternate between viewing a constraint formula ($\phi$)
% as a set of constraints and a conjunction of constraints in this
% description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
% \outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
% $(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
% conjunction of symmetric outlives constraints for the purpose of graph
% encoding. Conversely, given a digraph $G$, we define its constraint
% encoding $\Phi(G)$ in a straightforward manner. We say that a graph
% $G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
% \subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
% exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
% exist a path between same pair of vertices in $G_1$.

% A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
% \Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
% graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
% connected as $G(\phicstr)$. The problem of computing such a $G$ is
% equivalent to the problem finding of finding minimum number of edges
% to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
% Efficient algorithms to solve the later problem are known to
% exist~\cite{graph}, and we rely on them for the solution.

% Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
% graph augmentation problems, as described above, results in the
% following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
% outlives constraints are replaced with equalities):
% \begin{center}
% \(
%   \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
%      \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
%      \rho_2 = \rho_0 \conj \rho_3 = \rho_1
% \)\\
% \(
%   \varphi_2 \Leftrightarrow \rho_5 = \rho_0
% \)
% \end{center}
% Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
% $c_{14}$ gives us a recursive constraint of the following form:
% \begin{center}
% \(
%   \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
% \)
% \end{center}
% Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
% $F$ is a substitution function, not necessarily idempotent. Let
% $G(\varphi) = \phicstr \conj F(\varphi)$, so that the constraint is
% $\isvalid{\phictxt \conj \varphi}{H(\varphi)}$. Our solution to such
% recursive constraints  is based on some observations. Firstly, the set
% of all possible constraints over a finite set of region
% variables/identifiers is a complete meet semi-lattice, where (a).
% meet operation is defined as $\phi_1 \sqcap \phi_2 \triangleq \phi_1
% \conj \phi_2$, and (b). lattice order is defined as $\phi_1 \le \phi_2
% \triangleq \isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Secondly,
% function $H$ is a monotone over the lattice since $\forall
% \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow \phi_2)
% \Rightarrow (H(\phi_1) \Rightarrow H(\phi_2))}$. From Knaster-Tarski's
% theorem, we know that $H$ has a greatest fixed point $\phi_f$ such
% that $\isvalid{\cdot}{\phi_f \Leftrightarrow H(\phi_f) }$. To solve
% the recursive constraint shown above, we first compute the greatest
% fixed point ($\phi_f$) of $H$, and then reduce the constraint to the
% following non-recursive constraint:
% \begin{center}
% \(
%   \isvalid{\phictxt \conj \varphi}{\phi_f}
% \)
% \end{center}
% Which can then be solved via the graph-based approach described
% before.
